---
title: "hw4"
author: "Chenfeng Hao"
date: "`r format(Sys.time(), tz = 'EST', '%b %d, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = F, message = F, tidy = T)
library(rpart)
library(rpart.plot)
```

Part 1 – manual calculations

The table to the right contains information about patients in a study for a particular virus.  The attributes consist of id,gender, age group, pulse/heart rate, test and the class, virus. The test is the results of a medical test that can range from 1to 100.  It is thought that gender, ageGroup, pulse and/or test may be predictive of the presence of the virus.  Do the analysis to answer the questions below.

1. calculate the gini scores for splitting on  

a) gender 

**F: 1 - (2/6)^2^ - (4/6)^2^ $\approx$ 0.444  
M: 1 - (2/4)^2^ - (2/4)^2^ = 0.5  
gini_split~gender~ = 0.444 * 6/10 + 0.5 * 4/10 $\approx$ 0.466**

b) ageGroup (split by 1 & 2 on one side and 3 on the other) 

**1 & 2: 1 - (4/6)^2^ - (2/6)^2^ $\approx$ 0.444  
3: 1 - 0^2^ - 1^2^ = 0  
gini_split~ageGroup~ = 0.444 * 6/10 + 0 * 4/10 $\approx$ 0.266 **

2. calculate the gini scores for splitting on a)pulse  

**57: <= (1 - (1/1)^2^ = 0), > (1 - (6/9)^2^ - (3/9)^2^ $\approx$ 0.444), split = 0 * 1/10 + 0.444 * 9/10 $\approx$ 0.400  
60: <= (1 - (2/2)^2^ = 0), > (1 - (6/8)^2^ - (2/8)^2^ = 0.375), split = 0 * 2/10 + 0.375 * 8/10 = 0.3  
65: <= (1 - (3/3)^2^ = 0), > (1 - (6/7)^2^ - (1/7)^2^ $\approx$ 0.245), split = 0 * 3/10 + 0.245 * 7/10 $\approx$ 0.172  
67: <= (1 - (1/4)^2^ - (3/4)^2^ = 0.375), > (1 - (5/6)^2^ - (1/6)^2^ $\approx$ 0.278), split = 0.375 * 4/10 + 0.278 * 6/10 $\approx$ 0.317  
68: <= (1 - (2/5)^2^ - (3/5)^2^ = 0.48), > (1 - (4/5)^2^ - (1/5)^2^ = 0.32), split = 0.48 * 5/10 + 0.32 * 5/10 = 0.4  
70: <= (1 - (3/6)^2^ - (3/6)^2^ = 0.5), > (1 - (3/4)^2^ - (1/4)^2^ = 0.375), split = 0.5 * 6/10 + 0.375 * 4/10 = 0.45  
73: <= (1 - (4/7)^2^ - (3/7)^2^ $\approx$ 0.490), > (1 - (2/3)^2^ - (1/3)^2^ $\approx$ 0.444), split =  0.490 * 7/10 + 0.444 * 3/10 $\approx$ 0.476  
76: <= (1 - (5/8)^2^ - (3/8)^2^ $\approx$ 0.469), > (1 - (1/2)^2^ - (1/2)^2^ = 0.5), split = 0.469 * 8/10 + 0.5 * 2/10 $\approx$ 0.475  
79: <= (1 - (5/9)^2^ - (4/9)^2^ $\approx$ 0.494), > (1 - (1/1)^2^ = 0), split = 0.494 * 9/10 + 0 * 1/10 $\approx$ 0.445**
 
3. what is the best split of all the attributes? 

**The best split is pulse (pulse > 65), with a gini index of 0.172.**

4. after that attribute is selected, describe briefly in your own words, what the next step is.

**After choosing the best split of all attributes, the next step is to determine the next best attribute in terms of the split score, then the next best, and so on. The process can stop when all records belong to the same class or have similar attribute values. And this would finish building the the classification tree.**

Part 2 – classifying instances

In this part, we are using a different data set from the one above with more instances (thousands) but the same attributes. It is also not the same as the data set used in the next exercise below. You are not allowed to seethe data but you can see the decision tree that was built from the data set to the right.

Using just the tree, do the following: 

1.categorize the following instances: 

a) {2031, m, 1, 66, 50 ,?} 

**n**

b) {2032, f, 1, 90 ,81 ,?} 

**n**

c) {2033, m, 2, 75 ,45 ,?} 

**y**

2. characterize the patients with the virus by briefly describing their characteristics. 

**From the data, patients with the virus are all men, across all age groups; for younger men, patients with the virus have a test score of less than 50; for middle aged men, patients with the virus have a heart rate of over 60 and test score at least 50; for older men, patients with the virus have a heart rate of at most 60 and test score at least 50.**

3. if there was a middle aged male, with a test score > 50, what could he do to try to avoid the virus?

**Lower pulse/heart rate by reducing intensive physical activities or other means.**

Part 3 – building a model in R

You have been provided a data set that consists of data for the 100 US senators, as of summer 2020.  The attributes consist of name, political party (democrat, republican or independent), age, eight binary attributes for their previous career (law, military, social or non-profit, etc.), the number of years in office and finally, the class which is 0/1.  The class=0 means their voters don’t approve of them and class=1 means that the voters do approve of them.  This was taken from a website of senator rankings. 

Follow these directions to process the data, create models and examine the results and answer the questions: 

1. Perform the following actions 

a) read the data from the file into a data matrix 
\
```{r}
hw4data <- read.table("hw04dataSenate.txt", header = T)
```

b) use  rpart to build a decision tree model using approval as the class and all of the attributes except name 
\
```{r}
for (i in c(3, 5:12, 14)) {
  hw4data[, i] <- as.factor(hw4data[, i])
}

hw4data_noname <- hw4data[, 3:14]

mod_senate <- rpart(approval ~ ., hw4data_noname)
```

c) plot the decision tree using rpart.plot 
\
```{r}
rpart.plot(mod_senate)
```

2. What is one of the main determiners of voter approval? 

**Age, or more specifically (since two determiners related to age are being referenced), age >= 67, is the main determiner of voter approval, because, of the 41% data being covered, it correctly classifies 71% as having voter approval.**

3. Is a senator’s party important in predicting their approval rating?  Explain briefly. 

**From the data, a senator's party affiliation seems to make a difference to their approval rating, if they are less than 67 years of age and served at least 3 years in the seat. Being a republican does not make better approval rating more likely, but being a democrat has a 65% chance of getting approved by the voters.**
