---
title: "hw01"
author: "Chenfeng Hao"
date: "9/9/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1 

1. attrs: age, gender, lineage, title 

   class: characters who will get killed and those who will not

   Training set would be a subset of all characters in GOT used for fitting and and building the classification model.

2. attrs: programming experience, work experience, stats background, birth place, education background 
   
   proximity measures:\
     * programming experience -- years\
     * work experience -- years & job field/type\
     * stats background -- number of stats classes taken, levels of stats classes taken\
     * birth place -- country/state\
     * education background -- degree obtained/degree major\

# Part 2: paste in the command or results (as indicated) 

## 2. vectors 

a) create a vector named iq of 100 elements of N(100,20) (normal with a mean of 100 and std dev of 20) data.

```{r results='hide'}
iq <- rnorm(100, mean = 100, sd = 20)
```

b) add 10 to every element of iq

```{r results='hide'}
iq + 10
```

c) calculate the mean and std dev of iq

```{r}
mean(iq)
sd(iq)
```

d) display the first 10 elements of iq using an index (e.g. iq[1:10])

```{r results='hide'}
iq[1:10]
```

e) display the elements 11 through 20 of iq using an index

```{r results='hide'}
iq[11:20]
```

## 3. tables

a) create a matrix called tmp with the following values:

```{r}
matdata <- c(3, 7,-1, 6, 10, 0)
tmp <- matrix(matdata,
              nrow = 2,
              ncol = 3,
              byrow = TRUE)
```

b) read the file hw01data.txt in R and assign it to the variable x

```{r}
x <- read.table("data.txt", header = TRUE)
```

c) calculate the mean of column 3 

```{r}
mean(x[, 3])
```

d) sum up the values in column 2, rows 5 through 15 

```{r results='hide'}
sum(x[5:15, 2])
```

e) calculate the average rate of all states where the average income is > 90000, example below  
• idx=x[,3]>90000  
• mean(x[idx,2]) 

```{r}
ind <- x[, 3] > 90000
mean(x[ind, 2])
```

f) calculate the average avgInc for all states where the fico is < 700  
• using the method above (using idx)  
• using a for loop and if statement

```{r}
idx <- x[, 4] < 700
mean(x[idx, 3])

fmean <- 0
fcounter <- 0
ftotal <- 0
for (i in 1:nrow(x)) {
  if (x[i, 4] < 700) {
    ftotal <- ftotal + x[i, 3]
    fcounter <- fcounter + 1
  }
}
fmean <- ftotal / fcounter
fmean
```


4. functions  
a) write a function that takes a table and returns an vector that contains the mean values of columns 2, 3 and 4

```{r}
meanValuesOfColumns = function(tableName) {
  tableData <- read.table(tableName, header = TRUE)
  mean2 <- mean(tableData[, 2])
  mean3 <- mean(tableData[, 3])
  mean4 <- mean(tableData[, 4])
  meanValuesOfColumns <- c(mean2, mean3, mean4)
  return (meanValuesOfColumns)
}
```

