---
title: "hw02"
author: "Chenfeng Hao"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
library(dplyr)
```

Part A – concepts

1. there are many ways to handle missing data. For example, you can delete the attributes that have missing data, delete rows that have missing data, replace the data with a global average (average of the whole column/attribute) or replace the missing data with the class average (the average of column for only rows with the same class value). Suggest how you handle the following cases:

a) there are just a few instances with many missing attributes

**Remove the rows that have missing data.**

b) there are many instances (more than 50%) that have missing data for just one attribute

**There are too many missing values. It may be more appropriate to ignore the attribute in the analysis.**

c) about 10 percent of the instances have missing data in one homogeneous attribute.

**Use the mean of valid data values of the same attribute as an estimate for missing values.**

d) about 10 percent of the instances have missing data in one heterogeneous attribute (that is the values appear to follow different distributions depending on the class).

**Use the mean of valid data values of the same attribute class as an estimate for missing values.**

2. Under what circumstances would you remove an attribute?

**An attribute can be removed if there are too many missing values for that attribute.**

3. Assume we want to create a table with students as instances and attributes such as name, program, total credits and GPA. Also assume that you have a table with the student grades (one instance for each class, each student has taken). Describe how you would calculate the total credits and GPA for the student table from the grades table.

**Use data aggregation to combine data in the grade table into attributes that could be used for calculating total credits and GPA (credits for courses, weights, etc.).**

Part B – R coding

Follow the instructions below in R.

1. read in the data from the file hw02data.txt.
\
```{r}
hw02data <- read.table("hw02data.txt", header = TRUE)
head(hw02data, 1)
```

2. find the instances with missing values in column 2 and remove those instances.
\
```{r}
ind <- is.na(hw02data[,2])
hw02data_no_na <- hw02data[!ind,]
head(hw02data_no_na,1)
```

3. using an index vector, find all of the instances with a code of “USA” and assign them to a new table.
\
```{r}
index_usa <- hw02data_no_na[,2]=="USA"
hw02data_usa <- hw02data_no_na[index_usa,]
head(hw02data_usa,1)
```

4. create a 6th column and name it “infect”. Calculate the values to be the value from the 4th column of that instance minus the value of the 4th column (cases) of the prior instance (place zero in the first instance of column 6).
\
```{r}
# method 1
tmp <- hw02data_usa
col_4 <- tmp[, 4]
col_4_lag <- lag(col_4)
new_col <- col_4 - col_4_lag
table_with_6th_col_1 <- cbind(tmp, new_col)
colnames(table_with_6th_col_1)[6] <- "infect"
table_with_6th_col_1[1, 6] <- 0
head(table_with_6th_col_1,1)

# method 2
table_with_6th_col_2 <- cbind(hw02data_usa, 0)
colnames(table_with_6th_col_2)[6] <- "infect"
for (i in 2:nrow(table_with_6th_col_2)) {
  table_with_6th_col_2[i, 6] <-
    table_with_6th_col_2[i, 4] - table_with_6th_col_2[i - 1, 4]
}
head(table_with_6th_col_2,1)
```

5. create a new table with 1 row and 6 columns (entity, code, avg, max, days and growth). Replace entity and code with the values from the USA data. The rest should be zeros.
\
```{r}
usa_data <- table_with_6th_col_2
new_usa_data <- cbind(usa_data[1:2],0,0,0,0)
column_names <- c("entity", "code", "avg", "max", "days", "growth")
colnames(new_usa_data) <- column_names
head(new_usa_data,1)
```

6. replace avg with the mean of the cases for the USA data.
\
```{r}
new_usa_data[,"avg"] <- mean(usa_data[,"cases"])
head(new_usa_data,1)
```

7. replace max with the maximum of the cases for the USA data.
\
```{r}
new_usa_data[,"max"] <- max(usa_data[,"cases"])
head(new_usa_data,1)
```

8. replace days with the number of days that cases have been reported. In other words, count the number of days starting with the first non-zero value in the cases column.
\
```{r}
new_usa_data_0 <- new_usa_data
# method 1
cases_col <- usa_data[, "cases"]
transformed_cases_col <- ifelse(cases_col == 0, 0, 1)
new_usa_data[, "days"] <- cumsum(transformed_cases_col)
head(new_usa_data,1)

# method 2
temp <- vector()
for (i in 1:nrow(usa_data)) {
  if (usa_data[i, "cases"] == 0) {
    temp <- c(temp, 0)
  } else{
    temp <- c(temp, 1)
  }
}
new_usa_data_0[, "days"] <- cumsum(temp)
head(new_usa_data_0,1)

```

9. replace growth (column 6) with either a -1, 0 or 1. Calculate two mean values, one for the latest 10 days and the other for the 10 days prior to that.

a) if the first mean is less than the second one, enter -1.

b) if the first mean is more than 10% higher than the second, enter 1. 

c) otherwise, enter 0
\
```{r results='show'}
cases_daily_increase <- c(usa_data$cases[1])

for (i in 2:nrow(usa_data)) {
  cases_daily_increase <-
    c(cases_daily_increase, usa_data$cases[i] - usa_data$cases[i - 1])
}

for (i in 20:length(cases_daily_increase)) {
  if (mean(cases_daily_increase[i - 9]:i) < mean(cases_daily_increase[(i - 19):(i - 10)])) {
    new_usa_data[i, 6] <- -1
  } else if (mean(cases_daily_increase[i - 9]:i) > 1.1 * mean(cases_daily_increase[(i - 19):(i - 10)]))
  {
    new_usa_data[i, 6] <- 1
  } else{
    new_usa_data[i, 6] <- 0
  }
}

new_usa_data
```

