---
title: "hw9"
author: "Chenfeng Hao"
date: "`r format(Sys.time(), tz = 'EST', '%b %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dbscan)
```

part 1  

1. How many clusters do you think there are?  

**Answer: **  

2  

2. What would you consider points A and B?  

**Answer: **  

Outliers  

3. Do you think that A and B should be in the same cluster?  Why or why not?  

**Answer: **  

If A and B need to be placed in a cluster, they should be in the same cluster. However, I think the overall picture of the data points suggest they are more likely outliers.

part 2  

Read the data into R.  Find 2 clusters using kmeans (be sure to set.seed(1) before you cluster).  After that, use hierarchical, single link to find 2 clusters and then 3 clusters.  Do the same for complete link.  Finally use DBSCAN to find clusters.  Try it using eps=.35 and minPts=4 and then eps=.40 and minPts=4.  Answer the following questions (it will be helpful to include plots of the resulting clusters):  

```{r}
hw9.data <- read.table("data(1).txt", header = T)

distance <- dist(hw9.data)
```
(Assume A is the higher of the two data points and B the lower.)

1. For kmeans, which cluster did A belong to?  B?  Explain why.  

```{r}
set.seed(1)
km <- kmeans(hw9.data, 2)
plot(hw9.data, col=km$cluster)
```
\
As shown in the plot, A belongs to the left cluster and B the right, because, in terms of Euclidean distance (or chosen proximity measure), A is closer to the centroid of the left cluster and B that of the right.  

2. For single link with 3 clusters, where did it place A and B?  Explain why.  

```{r}
sng <- hclust(distance, method = "single")

sng3 <- cutree(sng, 3)
plot(hw9.data, col=sng3)
```
\
As shown in the plot, A and B are in the same cluster, separate from the other two clusters. This is because single link uses MIN as the proximity measure and points A, B are obviously further enough from the closest points of the other two clusters to be able to form their own cluster.  

3. For single link with 2 clusters, where did it place A and B?  Explain why.  

```{r}
sng2 <- cutree(sng, 2)
plot(hw9.data, col=sng2)
```
\
As shown in the plot, both A and B are in the left plot, because they are closer to the closest points in the left cluster than that in the right.

4. For complete link, explain how the algorithm went from 3 clusters to 2.  

```{r}
compl <- hclust(distance, method = "complete")

compl2 <- cutree(compl, 2)
compl3 <- cutree(compl, 3)

par(mfrow = c(1, 2))
plot(hw9.data, col = compl3, main = "3 clusters, complete")
plot(hw9.data, col = compl2, main = "2 clusters, complete")
```
\
Complete uses max as the proximity measure. The distance from A (and B) to the furthermost point in the left cluster is closer than that between A (and B) and the furthermost point in the left cluster, so both A and B are assigned to the left cluster.  

5. For DBSCAN, how many noise points were there when eps=.35?  How many when eps=.40?  Explain the difference.  If you kept eps=.40 what could you change minPts to so that A and B were no longer noise points.  

```{r}
db1 <- dbscan(hw9.data, eps = .35, minPts = 4)
db1
plot(hw9.data, col=db1$cluster)


db2 <- dbscan(hw9.data, eps = .40, minPts = 4)
db2
plot(hw9.data, col=db2$cluster)

db3 <- dbscan(hw9.data, eps = .40, minPts = 2)
db3
plot(hw9.data, col=db3$cluster)
```
\
eps = .35, minPts = 4, noise points = 13  
eps = .40, minPts = 4, noise points = 2  
eps = .40, minPts = 2, noise points = 0